### 2022-06-13 paper.tex
100kb window -> 100kb tile

\begin{minipage}{0.7\textwidth}
\small
\textit{Abstract} -- Single-cell-bisulfite sequencing (scBS) allows to asses DNA methylation with single-basepair and single-cell resolution. Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability. Currently, this is simply done by tiling the genome into large windows and averaging in these. We show that this coarse-graining dilutes signal, and suggest improved approaches to determine informative regions to quantify for and to quantitate methylation in these windows. We demonstrate how our preprocessing approach helps to better distinguish cell types, reduces demands on cell number, and show that our method of finding variably-methylated regions finds biologically meaningful loci. Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality requried to work with scBS data.
\end{minipage}
\vspace{6.5ex}

@@

This procedure is commonly adapted when working with single-cell DNA-methylation data, so that established methods used in scRNA-Seq can be used.
The task is therefore how to construct a suitable matrix as input for the PCA. A simple and common approach, used for instance by \citet{luo2017single}, is to tile the genome into windows of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each window. To this end, we identify all CpG sites in the window for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell. Then we denote as average DNA methylation of this window in a given cell the proportion of observed CpG sites in the window that were found to be methylated (Figure \ref{fig:smoothres}A). This yields a matrix, with one row for each cell and one column for each window from the genome tiling, comprising numbers (methylation fractions) between 0 and 1. This matrix is now subjected to PCA. After the PCA, one can proceed using approaches known from scRNA-seq.

@@

The standard PCA algorithm cannot deal with missing data. Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most windows and discard any interval that is not covered by any single cell. Hence, a naive analysis has to resort to a coarse tiling of the genome. To counter this, we propose a simple and straightforward way to deal with missing data in the input matrix to a PCA: In a first iteration, we replace each missing value in the centered input matrix with zero, then run the PCA. Then, only these zeroes are replaced by the value predicted by the PCA and the PCA is rerun. Convergence is typically achieved after two to five iterations. For details, see Methods.

@@

In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each such window. Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.

@@

\begin{figure*}
	\begin{center}
		A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
		\hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
		\hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
		C\hspace*{.7\textwidth}~\\
		\hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
		D\hspace*{.7\textwidth}~\\
		\hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
		\caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}. \textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic regions (VMRs or 100~kb windows) as input to PCA. Exemplary UMAPs for encircled analyses are shown in (C). \textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells. Exemplary UMAPs for analyses are shown in (D). \textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or mean shrunken residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100~kb windows (left) or mean shrunken residuals of VMRs (right). Data from \citet{luo2017single}.}
		\label{figure:score}
	\end{center}
\end{figure*}

To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}. This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
We repeated this analysis several times, either using 100~kb windows or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation. 
For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 high-coverage regions that were selected for PCA.
Specifically, we ranked regions by their number of coverging cells (i.e., cells with at least on read in the region) and selected those with the highest number.

@@

Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
In contrast, the neighbor count score for the top 1000 genomic 100~kb windows is much lower.
To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
% why not just plot PC1 vs. PC2 here? % because people will want to see the UMAP. -S
Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{figure:score}D).

@@

\subsection{Re-analysis of published scBS datasets}
To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb windows, or by acquiring VMRs with \texttt{scbs scan} using default options. For all regions (VMRs and 100~kb windows), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}. Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded. To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in. 
For dimensionality reduction, missing values were imputed with an iterative PCA as described above. To visualize data, UMAP was performed on the final 15 principal components.

>>>>

\begin{minipage}{0.7\textwidth}
\small
\textit{Abstract} -- Single-cell-bisulfite sequencing (scBS) allows to asses DNA methylation with single-basepair and single-cell resolution. Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability. Currently, this is simply done by dividing the genome into large tiles and averaging in these. We show that this coarse-graining dilutes signal, and suggest improved approaches to determine informative regions to quantify for and to quantitate methylation in these tiles. We demonstrate how our preprocessing approach helps to better distinguish cell types, reduces demands on cell number, and show that our method of finding variably-methylated regions finds biologically meaningful loci. Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality requried to work with scBS data.
\end{minipage}
\vspace{6.5ex}

@@

This procedure is commonly adapted when working with single-cell DNA-methylation data, so that established methods used in scRNA-Seq can be used.
The task is therefore how to construct a suitable matrix as input for the PCA. A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each tile. To this end, we identify all CpG sites in the tile for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell. Then we denote as average DNA methylation of this tile in a given cell the proportion of observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A). This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (methylation fractions) between 0 and 1. This matrix is now subjected to PCA. After the PCA, one can proceed using approaches known from scRNA-seq.

@@

The standard PCA algorithm cannot deal with missing data. Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most tiles and discard any interval that is not covered by any single cell. Hence, a naive analysis has to resort to a coarse tiling of the genome. To counter this, we propose a simple and straightforward way to deal with missing data in the input matrix to a PCA: In a first iteration, we replace each missing value in the centered input matrix with zero, then run the PCA. Then, only these zeroes are replaced by the value predicted by the PCA and the PCA is rerun. Convergence is typically achieved after two to five iterations. For details, see Methods.

@@

In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each such tiles. Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.

@@

\begin{figure*}
	\begin{center}
		A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
		\hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
		\hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
		C\hspace*{.7\textwidth}~\\
		\hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
		D\hspace*{.7\textwidth}~\\
		\hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
		\caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}. \textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic regions (VMRs or 100~kb tiles) as input to PCA. Exemplary UMAPs for encircled analyses are shown in (C). \textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells. Exemplary UMAPs for analyses are shown in (D). \textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb tiles (left) or mean shrunken residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100~kb tiles (left) or mean shrunken residuals of VMRs (right). Data from \citet{luo2017single}.}
		\label{figure:score}
	\end{center}
\end{figure*}

To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}. This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation. 
For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 high-coverage regions that were selected for PCA.
Specifically, we ranked regions by their number of coverging cells (i.e., cells with at least on read in the region) and selected those with the highest number.

@@

Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
In contrast, the neighbor count score for the top 1000 genomic 100~kb tiles is much lower.
To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
% why not just plot PC1 vs. PC2 here? % because people will want to see the UMAP. -S
Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{figure:score}D).

@@

\subsection{Re-analysis of published scBS datasets}
To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using default options. For all regions (VMRs and 100~kb tiles), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}. Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded. To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in. 
For dimensionality reduction, missing values were imputed with an iterative PCA as described above. To visualize data, UMAP was performed on the final 15 principal components.
